---
title: "Motor Trend Assignment"
author: "Alan C Bonnici"
date: "Tuesday, October 21, 2014"
output: pdf_document
---

#Assignment

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:  
* "Is an automatic or manual transmission better for MPG"  
* "Quantify the MPG difference between automatic and manual transmissions"


## Data

The data source for this assignment comes from the **mtcars** dataset which is bundled with R. The dataset can be loaded into R by typing `r data(mtcars)`. The effect of this command is to load into the environment **mtcars**. The resulting dataset consists of **`r nrow(mtcars)`** rows, each row made up of **`r ncol(mtcars)`** columns. The information about this data can be called up by typing *?mtcars* at the R prompt. A link to the online version of the page is available at (https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).

The propoerty *am* (indicating wether the transmission is manual or automatic) will be converted into a factor variable.

```{r}
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <-c("Automatic", "Manual")
```

The boxplot (Fig 1) explains the different transmission types with respect to the mpg. From this chart, it seems that manual transmission have a higher mpg than automatic transmissions. One could also not that neither have any outliers and the whiskers of the box plot are balanced.

The histograms (Fig 2) compare the mpg of the two types of transmissions.

```{r}
cars.Auto <- mtcars[mtcars$am == "Automatic",]
cars.Man <- mtcars[mtcars$am == "Manual",]
```

The average mpg for vehicles with automatic transmission is **`r round(mean(cars.Auto$mpg),2)`**. The average mpg for vehicles with manual transmission is **`r round(mean(cars.Man$mpg), 2)`**.

```{r}
t.test(cars.Auto$mpg, cars.Man$mpg)
```

From the information presented in the Welch Two Sample t-test, it is clear that at a 95% confidence interval the difference between the mpg means of the automatic and petrol vehicles is always negative implying that manual will always be 3.21 mpg more efficient than automatic.

From the information presented above, the *null hypothesis: true difference in means is equal to 0* will be rejected at the 99% confidence level.

Even though the information shown above indicates that manual is more efficient than automatic, one needs to ensure that other factors are not confounding the results. For example, if automatic cars analysed here are heavier than manual ones than weight could be "upsetting" the results. Likewise if one category of vehicle have higher, hp or cylinder values this could be a factor and not the transmission.

A correlation analysis can help show the relationship of the various properties with respect to mpg shows that an increase in cylinders, displacement, hp, weight and number of carburetors all negativly impact mpg while rear axle ratio, 1/4 mile time and number of forward gears all contribute to higher mpgs.

```{r echo=FALSE}
x <- head(cor(sapply(mtcars, as.numeric)),1)
x <- x[, order(x)]
x[-11]
```

In a simple regression analysis from mpg given transmission 
```{r echo=FALSE}
fit <- lm(mpg ~ am, data = mtcars)
summary(fit)
```

From the above data we can infer that, on average, manual transmission vehicle give 7.25 mpgs more than then automatic transmission. A value of R^2 of 35.98% means that the model does not explain all of the variance and further investigation is required.

In order to investigate further a multivariate regression analysis is performed. In this case we adopt a stepwise linear regression. Stepwise regression is a semi-automated process of building a model by successively removing variables based solely on the t-statistics of their estimated coefficients. 

```{r echo=FALSE}
fit2 <- lm(mpg ~ ., data=mtcars)
summary(fit2)

fit2.best <- step(fit2, trace=0)
summary(fit2.best)
```

From the information presented above we can conclude that manual transmission contributes 2.94 mpg over automatic. This is far less than the value reported using the linear regression of mpg given transmission on its own or the t-test and can be explained by the weight and 0.25 mile time that play an important role in overall mpg value. See Fig 3 are a series of charts relating to the step wise regression.

### Fig 1
```{r echo=FALSE , fig.height=4}
boxplot(mpg~am, data = mtcars,
        xlab = "Transmission Type",
        ylab = "mpg",
        main = "mpg by Transmission Type")
```

### Fig 2
```{r fig.height=4, echo=FALSE}
par(mfrow=c(1,2))
bins <- seq(10, 35, by=2.5)
hist(mtcars$mpg[mtcars$am == "Automatic"], breaks=bins, main="mpg for automatic vehicles", xlab="mpg", xlim=c(10, 35))

hist(mtcars$mpg[mtcars$am == "Manual"], breaks=bins, main="mpg for manual vehicles", xlab="mpg", xlim=c(10, 35))
par(mfrow=c(1,1))

```

### Fig 3
```{r fig.height=8, echo=FALSE}
par(mfrow = c(2,2))
plot(fit2.best)
par(mfrow = c(1,1))
```
