---
title: "Motor Trend Assignment"
author: "Alan C Bonnici"
date: "Tuesday, October 21, 2014"
output: pdf_document
---

#Assignment

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions"


## Data

The data source for this assignment comes from the **mtcars** dataset which is bundled with R.

The dataset can be loaded into R by typing `{r data(mtcars)}`. The effect of this command is to load into the environment **mtcars**.

The resulting dataset consists of **`{r nrow(mtcar)}`** rows, each row made up of **`{r ncol(mtcar)}`** colunms.

The columns making up this dataset are:

1 mpg	 Miles/(US) gallon  
2 cyl	 Number of cylinders  
3 disp	 Displacement (cu.in.)  
4 hp	 Gross horsepower  
5 drat	 Rear axle ratio  
6 wt	 Weight (lb/1000)  
7 qsec	 1/4 mile time  
8 vs	 V/S  
9 am	 Transmission (0 = automatic, 1 = manual)  
10 gear	 Number of forward gears  
11 carb	 Number of carburetors  

The information provided here was sourced from the R documentation page. This, together with addition information can be called up by typing *?mtcars* at the R prompt. A link to the online version of the page is available at (https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).

Below is a sample of data:

```{r}
tail(mtcars)
```


# Project Requirement: **Is an automatic or manual transmission better for MPG**

The two columns we will be dealing with are *mpg* and *am*. The data type of these two properties are:

* mpg    `{r class(mtcars$mpg)}'  
* am     `{r class(mtcars$am)}'

*am* will be converted into a factor variable.

```{r}
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <-c("Automatic", "Manual")
```

```{r}
boxplot(mpg~am, data = mtcars,
        xlab = "Transmission Type",
        ylab = "mpg",
        main = "mpg by Transmission Type")
```

The boxplot above examples the different transmission types with respect to the mpg. From this chart, it seems that manual transmission have a higher mpg than automatic transmissions. One could also not that neither have any outliers and the wiskers of the box plot are balanced.

```{r fig.align='center', echo=TRUE, fig.height=8}
par(mfrow=c(2,1))
hist(mtcars$mpg[mtcars$am=="Automatic"], breaks=12, main="mpg for automatic vehicles", xlab="mpg", xlim=c(10, 35))
hist(mtcars$mpg[mtcars$am=="Manual"], breaks=12, main="mpg for manual vehicles", xlab="mpg", xlim=c(10,35))
par(mfrow=c(1,1))
```

